---
title: '**_PREDICTING OPPORTUNITY VALUES FOR SMEs_** - Capstone Project of HarvardX
  _Data Science Professional Certificate Course_ - Harvard University'
author: "Miguel Ángel Boto Bravo, PhD in Linguistics"
date: "16/09/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)

```


# *1. INTRODUCTION*

Predicting many of the sales variables involved in the preparation of an annual forecast in small, and / or recently created companies, is a major problem that affects the strategic planning of SME companies. Among the main variables, associated or not with competition in the market, we must highlight a low volume of historical data and the absence of data due to insufficient data collection.

The objective of the project that we present is to analyze the effectiveness of the application of some algorithm models in the prediction of sales values and, therefore, their use in the preparation of annual forecasts.

For this, we will use the data of sales opportunities of new clients of a small software company that contains, as we will see in the exploratory data analysis, limitations that affect the volume of data collected and the characteristics that make up the database.

# *2. EXPLORATORY ANALYSIS AND DATA VISUALIZATION.*

## *2.1 Installing libraries and preparing the database:*

Before starting the exploratory analysis of Sales Report, we must install and run the following libraries:

```{r}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("readxl", repos = "http://cran.us.r-project.org")
if(!require(readxl)) install.packages("readxl", repos = "http://cran.us.r-project.org")
if(!require(openxlsx)) install.packages("openxlsx", repos = "http://cran.us.r-project.org")


library(tidyverse)
library(caret)
library(data.table)
library(lubridate)
library(readxl)
library(openxlsx)

```

Next, we will download the Sales Report dataset with which we are going to work, format the dataset and create a training and validation sets:


```{r}
#Import dataset

url <-"https://github.com/miguelboto/sales_report/blob/master/SR.xlsx?raw=true"

download.file(url, destfile = "SR.xlsx", mod="wb")

data <- read.xlsx("SR.xlsx")

#Converting date numeric rows to date format and including a new row with time in days (difference between open date and close date data):

sales_report <- data %>% mutate(open_date = as.Date(open_date, origin = "1899-12-30"), 
                close_date = as.Date(close_date, origin = "1899-12-30"), 
                last_date = as.Date(last_date, origin = "1899-12-30"),
                time = close_date - open_date)


# Validation: 10% of Sales Report data
set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`

test_index <- createDataPartition(y = sales_report$price, times = 1, p = 0.1, list = FALSE)
edx <- sales_report[-test_index,]
temp <- sales_report[test_index,]

# Opportunity_ID and count_id in validation set must be also in temp set

validation <- temp %>% 
  semi_join(edx, by = "opportunity_name") %>%
  semi_join(edx, by = "count_id")

removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

```

## *2.2 Exploratory analysis:*

As we can see below, the database is composed of 730 rows and 12 columns:

```{r}
dim(sales_report)


head(sales_report)


```

Each line corresponds to a business opportunity with new potential clients. Since it is a provision of services under the SaaS model, it is not a sale of the finished product upon delivery, but the value expressed in the _price_ column indicates the annual turnover of each new customer gained or, failing that, the potency of each lost customer.

Therefore, the key variable for our analysis is in the _price_ field, which shows the annual value in $ of each opportunity.

Likewise, the _status_ variable indicates if the opportunity was _won_ or _lost_, _open_date_ the date of opening the negotiations, _close_date_ the date of the end of the opening and _time_ the time elapsed between the opening and closing.

For its part, we can use two more variables of interest for the analysis: the origin of the potential customer (_sale_class_), that is, the sales channel through which the business opportunity was generated, and the type of customer (_market_class_).

The _sale_class_ category contains 7 variables: website, self generated, campaign, word of mouth, bidding, partner and trade show. For its part, _Market_class_, has 5 variables: H (hospitals), CE (clinical engineering), O (others), M (manufacturers), TA (technical assistance), which correspond to the five markets that our software company can cater with your product.

Next we show the summary of the data contained in the dataset:

```{r}
summary(sales_report)
```

The data was collected between February 2014 (the start date of the company's activities) and September 2020. In the first plot we can see the distribution of opportunities by time, value and status. In the second one, the distribution per origin and status:

```{r}
#Opportunities value per time

sales_report %>% mutate(time = as.numeric(sales_report$time)) %>% 
  ggplot(aes(time, price, color = status)) + 
  geom_point(aes(shape = market_class)) +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10")



```

```{r}
sales_report %>% group_by(open_date, sale_class)  %>% 
  ggplot(aes(open_date, price, color = sale_class)) +
  geom_point() +
  scale_y_continuous(trans = "log10") +
  facet_grid(~status)
```


The difference between the value of the minimum and the maximum opportunity is striking, which shows that the provision of services offered by the company serves both small companies and large corporations. In the following boxplot we can see the distribution in greater detail:

```{r}
#Histogram Price Distribution All Opportunities per status
sales_report %>% 
  ggplot(aes(price)) +
  geom_histogram(bins = 30, fill = "pink", col = "black") +
  ggtitle("Price Distribution per Opportunity Status") +
  facet_grid(~status) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

However, we can see that there is a greater volume of opportunities concentrated in lower bands. Below we see the average price per opportunity (all Opportunities) and the average of won opportunities and success rate:

```{r }

#Average Price per status opportunity and Success Rate

dat <- sales_report %>% group_by(status) %>% 
  summarize(Opps = n(), Ticket = mean(price))


dat %>% mutate(Rate = (dat[, 2:2]) / sum(dat[, 2])) %>% knitr::kable()



```

```{r}
#Resume Prospects Opportunities 

sales_resume <- sales_report %>% 
  summarize(
    'Counts'=n_distinct(count_id),
    'Opportunities'=n_distinct(opportunity_ID),
    'Min_Price'=min(price),
    'Max_Price'=max(price), 
    'Average_Price'=mean(price),
    'Total_Values'=sum(price)
  ) 
sales_resume %>% knitr::kable()

sales_resume_won <- sales_report %>%  filter(status ==  "Won") %>% 
  summarize(
    'Counts'=n_distinct(count_id),
    'Won'=n_distinct(opportunity_ID),
    'Min Price'=min(price),
    'Max Price'=max(price), 
    'Average Price'=mean(price),
    'Total Values Won'=sum(price),
    'Success Rate'=(Won/sales_resume$Opportunities)
  ) 
sales_resume_won %>% knitr::kable()

sales <- sales_report %>% group_by(sale_class) %>%
  summarize(totalopp = n(), averageprice = mean(price)) %>%
  arrange(desc(averageprice))
sales %>% mutate(sale_class = reorder(sale_class, totalopp)) %>%
  ggplot(aes(x= sale_class, y = averageprice, color = sale_class)) +
  geom_point(aes(size=totalopp)) +
  ggtitle("Medium Ticket per Sale Class") +
  xlab("Sale Class")+
  ylab("Medium Ticket") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  guides(size=FALSE)

```

Continuing with the analysis of the dataset based on the price of the opportunities, the following boxplot shows us the distribution by value and type of market of potential clients, in which we can perceive that the largest volume is concentrated in hospitals and technical assistance:

```{r}
#Boxplot Opportunities and prices distribution per Market Class
sales_report %>% 
  group_by(market_class) %>% arrange(desc(price)) %>%
  ggplot(aes(market_class, price)) +
  scale_y_continuous(trans = "log10") +
  geom_boxplot(coef=3) +
  geom_jitter(width = 0.1, alpha = 0.2)  +
  ggtitle("Opportunities and Prices distribution per Market Class and Status")

type <- sales_report %>% group_by(market_class) %>%
  summarize(totalopp = n(), averageprice = mean(price)) %>%
  arrange(desc(averageprice))
type %>% mutate(market_class = reorder(market_class, totalopp)) %>%
  ggplot(aes(x= market_class, y = averageprice, color = market_class)) +
  geom_point(aes(size=totalopp)) +
  ggtitle("Medium Ticket per Market Class") +
  xlab("Market Class")+
  ylab("Medium Ticket") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  guides(size=FALSE)
```


The following histograms will help us to better understand the irregular distribution of the value of opportunities taking into account several variables:

```{r}

#Histogram Price Distribution per Market Class
sales_report %>% 
  ggplot(aes(price)) +
  geom_histogram(bins = 30, fill = "orange", col = "black") +
  ggtitle("Price distribution per Market Class (All Opportunities)") +
  facet_grid(~market_class) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

#Boxplot Opportunities per Market Class and Status
sales_report %>% 
  group_by(status) %>% arrange(desc(price)) %>%
  ggplot(aes(status, price, color = status)) +
  scale_y_continuous(trans = "log10") +
  geom_boxplot(coef=3) +
  geom_jitter(width = 0.1, alpha = 0.2) +
  facet_grid(~market_class) +
  ylab("Oportunities per Market Class ") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


#Histogram Price Won Distribution per Market Class
sales_report %>% filter(status == "Won") %>%
  ggplot(aes(price)) +
  geom_histogram(bins = 30, fill = "green", col = "black") +
  ggtitle("Price distribution of Opportunities Won per Market Class") +
  facet_grid(~market_class) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

#Histogram Won Price Distribution per vendor 
sales_report %>% filter(status != "won") %>%
  ggplot(aes(price)) +
  geom_histogram(bins = 30, fill = "yellow", col = "black") +
  ggtitle("Price distribution of Won Opportunities per vendor") +
  facet_grid(~vendor) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

#Histogram Lost Price Distribution per vendor 
sales_report %>% filter(status != "Lost") %>%
  ggplot(aes(price)) +
  geom_histogram(bins = 30, fill = "red", col = "black") +
  ggtitle("Price distribution of Lost Opportunities per vendor") +
  facet_grid(~vendor) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```

Finally, let's see the accuracy and standard deviation data before entering the predictions and results section:

```{r}
#Overall accuracy

y_hat <- sample(c("Won", "Lost"), length(sales_report), replace = TRUE) 

accuracy <- mean(y_hat == sales_report$status)

accuracy

#Price Mean & SD 

sales_report %>% group_by(status) %>% summarize(mean(price), sd(price)) 
```

# *3. PREDICTIONS AND RESULTS*

Bearing in mind that the general approach to defining the best in machine learning is to define a _ loss function_, we will make our prediction testing seven generative models with RMSE (the most widely used loss function) as a measure to evaluate our algorithm performance until reaching a RMSE value equal to or less than 17451, since this is the average value of the opportunities won over the 6 years analyzed in the dataset.

## *3.1 Naive By Mean Method*


```{r}
#Naive by Mean Method
mu <- mean(edx$price)
naive_RMSE <- RMSE(temp$price, mu) 

RMSE_results = data.frame(Method = "Naive Model", RMSE = naive_RMSE) 
RMSE_results


```
## *3.2 Time Effect Model*

```{r}

#Time effect model#
time_avg <- edx %>% group_by(time) %>%
  summarise(ti = mean(price - mu))

qplot(ti, data = time_avg, bins = 10, color = I("blue"))


prediction1 <- mu + temp %>% 
  left_join(time_avg, by="time") %>%
  pull(ti)

RMSE_time_effect_model <- RMSE(prediction1, temp$price)

RMSE_time_effect_model

RMSE_results <-  bind_rows(RMSE_results, 
                           data.frame(Method = "Time Effect Model", 
                                      RMSE = RMSE_time_effect_model))
RMSE_results %>% knitr::kable()


```

## *3.3 Time and Sale Class Effect Model*

```{r}
#Time and Sale Class effect model

sale_avg <- edx %>%
  left_join(time_avg, by='time') %>%
  group_by(sale_class) %>%
  summarize(sc = mean(price - mu - ti))

qplot(sc, data = sale_avg, bins = 10, color = I("blue"))

prediction2 <- temp %>%
  left_join(time_avg, by='time') %>%
  left_join(sale_avg, by='sale_class') %>%
  mutate(pred = mu + ti + sc) %>%
  pull(pred)
RMSE(prediction2, temp$price)

RMSE_time_and_sale_class_effect_model <- RMSE(prediction2, temp$price)

RMSE_results <- bind_rows(RMSE_results, 
                          data.frame(Method="Time & Sale Class Effect Model", 
                                     RMSE = RMSE_time_and_sale_class_effect_model))
RMSE_results %>% knitr::kable()


```

## *3.4 Time, Sale and Market Class Effect Model*

```{r}
#Time, Sale and Market Class Effect model

market_avg <- edx %>%
  left_join(time_avg, by='time') %>%
  left_join(sale_avg, by='sale_class') %>%
  group_by(market_class) %>%
  summarize(ma = mean(price - mu - ti - sc))


prediction3 <- temp %>%
  left_join(time_avg, by='time') %>%
  left_join(sale_avg, by='sale_class') %>%
  left_join(market_avg, by="market_class") %>%
  mutate(pred = mu + ti + sc + ma) %>%
  pull(pred)
RMSE(prediction3, temp$price)

RMSE_time_and_sale_and_market_class_effect_model <- RMSE(prediction3, temp$price)

RMSE_results <- bind_rows(RMSE_results, 
                          data.frame(Method="Time, Sale & Market Class Effect Model", 
                                     RMSE = RMSE_time_and_sale_and_market_class_effect_model))
RMSE_results %>% knitr::kable()


```

Surprisingly, including the market class variable has not improved the RMSE of the previous algorithm. Let's see the effect of including the Status variable.

## *3.5 Time, Sale, Market Class and Status Effect Model*

```{r}

#Time, Sale & Market Class and Status Effect Model



status_avg <- edx %>%
  left_join(time_avg, by='time') %>%
  left_join(sale_avg, by='sale_class') %>%
  left_join(market_avg, by='market_class') %>%
  group_by(status) %>%
  summarize(st = mean(price - mu - ti - sc - ma))

prediction4 <- temp %>%
  left_join(time_avg, by='time') %>%
  left_join(sale_avg, by='sale_class') %>%
  left_join(market_avg, by="market_class") %>%
  left_join(status_avg, by="status") %>%
  mutate(pred = mu + ti + sc + ma + st) %>%
  pull(pred)
RMSE(prediction4, temp$price)

RMSE_time_sale_market_class_and_status_effect_model <- RMSE(prediction4, temp$price)

RMSE_results <- bind_rows(RMSE_results, 
                          data.frame(Method="Time, Sale, Market Class & Status Effect Model", 
                                     RMSE = RMSE_time_sale_market_class_and_status_effect_model))
RMSE_results %>% knitr::kable()




```

Including the Status variable also has worsened the result obtained not by the previous model, but the third one (the best so far). We will include below the last of the variables, Vendor.

## *3.6 Time, Sale, Market Class, Status and Vendor Effect Model*

```{r}
#Time, Vendor, Sale & Market Class and Status Effect Model

vendor_avg <- edx %>%
  left_join(time_avg, by='time') %>%
  left_join(sale_avg, by='sale_class') %>%
  left_join(market_avg, by='market_class') %>%
  left_join(status_avg, by="status") %>%
  group_by(vendor) %>%
  summarize(ven = mean(price - mu - ti - sc - ma - st))


prediction5 <- temp %>%
  left_join(time_avg, by='time') %>%
  left_join(sale_avg, by='sale_class') %>%
  left_join(market_avg, by="market_class") %>%
  left_join(status_avg, by="status") %>%
  left_join(vendor_avg, by="vendor") %>%
  mutate(pred = mu + ti + sc + ma + st + ven) %>%
  pull(pred)
RMSE(prediction5, temp$price)

RMSE_time_sale_market_class_vendor_and_status_effect_model <- RMSE(prediction5, temp$price)

RMSE_results <- bind_rows(RMSE_results, 
                          data.frame(Method="Time, Sale, Market Class, Status & Vendor Effect Model", 
                                     RMSE = RMSE_time_sale_market_class_vendor_and_status_effect_model))
RMSE_results %>% knitr::kable()


```

Including the Vendor variable along with the rest of the variables tested has resulted in obtaining the lowest value of RSME.

Finally, let's see if we can improve the result by applying the Regularization process.

*3.6 Regularization of Time, Sale, Market Class, Status and Vendor Effect Model*

```{r}

#Regularization


lambdas <- seq(-0, 10, 0.25)
RMSES <- sapply(lambdas, function(l){
  mu <- mean(edx$price)
  ti <- edx %>%
    group_by(time) %>%
    summarize(ti = sum(price - mu)/(n()+l))
  sc <- edx %>%
    left_join(ti, by="time") %>%
    group_by(sale_class) %>%
    summarize(sc = sum(price - ti - mu)/(n()+l))
  ma <- edx %>%
    left_join(ti, by="time") %>%
    left_join(sc, by="sale_class") %>%
    group_by(market_class) %>%
    summarize(ma = sum(price - ti - sc - mu)/(n()+l))
  st <- edx %>%
    left_join(ti, by="time") %>%
    left_join(sc, by="sale_class") %>%
    left_join(ma, by="market_class") %>%
    group_by(status) %>%
    summarize(st = sum(price - ti - sc - ma - mu)/(n()+l))
  ven <- edx %>%
    left_join(ti, by="time") %>%
    left_join(sc, by="sale_class") %>%
    left_join(ma, by="market_class") %>%
    left_join(st, by="status") %>%
    group_by(vendor) %>%
    summarize(ven = sum(price - ti - sc - ma - st - mu)/(n()+l))
  
  predicted_prices <- temp %>%
    left_join(ti, by="time") %>%
    left_join(sc, by="sale_class") %>%
    left_join(ma, by="market_class") %>%
    left_join(st, by="status") %>%
    left_join(ven, by="vendor") %>%
    mutate(pred = ti + sc + ma + st + ven + mu) %>%
    pull(pred) 
  return(RMSE(predicted_prices, temp$price))
})

qplot(lambdas, RMSES)

lambda <- lambdas[which.min(RMSES)]
lambda

RMSE_results <- bind_rows(RMSE_results, 
                          data.frame(Method="Regularized Status, Market, Sale Class, 
                                     Vendor & Time Effect Model", RMSE = min(RMSES)))
RMSE_results %>% knitr::kable()

#Boxplot comparative effect models 

RMSE_results %>%  mutate(Method = reorder(Method, RMSE)) %>% 
  ggplot() + 
  geom_boxplot(aes(Method, RMSE, , size= RMSE)) + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# *4. CONCLUSION*

We have achieved the lowest RSME result, $13,380.32, with the Time, Sale, Market Class, Status & Vendor Effect Model. The Regularization model has not improved the result of this one.

Therefore, we can conclude that the expected opportunity value is $13,380.32, for which we will have to take into account all the variables available in our dataset, namely, Time, Origin of the Client, Market & Sale class and Status. 